Tail Flexibility in the Degrees of Preferential

### 2025-05-21 paper.qmd
Updating draft

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    fig-pos: 'H'
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, extremes]
bibliography: refs.bib

---
```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(paletteer)
library(gridExtra)
library(latex2exp)
library(ggpubr)
library(Rcpp)
library(RcppArmadillo)
```

# Introduction {#sec-intro}

Networks appear in many fields such as sociology, politics, epidemiology, and economics. Statistical methods have been and continue to be used to study networks that appear in these fields and beyond, from the use of stochastic block models to detect communities within the French political blogosphere [@Latouche11], to using Exponential Random Graph Models (ERGMs) to analyse the global trade network [@Setayesh22], and the use of mechanistic models in neuroscience to investigate the patterns of connections in neural systems [@Betzel17]. One such mechanistic model is the general preferential attachment (PA) model where as a network grows and new vertices join they connect to those already present in the network with weights proportional to some preference function $b$ of their degree. The well-known Barabási-Albert (BA) model is a special case of this when $b\sim k$ [BA99].

Mechanistic models, like the PA model, are in general only particularly useful when simulating networks and not fitting to the real data as usually only snapshots of real networks are available and not the entire evolution. As such, one of the most informative statistics when regarding a network is the degree distribution. When looking at the degree distributions of real networks it is attractive to many to use the power law to model the degree distribution, as the BA model generates degrees that follow a power law, and therefore claim that since the degree distribution is roughly power law then the underlying growth mechanism may be preferential attachment such as the BA model.  However, it has been shown that a general preference function will not always lead to a power law degree distribution [@krapivsky01] and additionally it has been debated for decades whether most real data actually follows a power law [@Voitalov_2019] or not[@Broido_2019].

Throughout this debate the use of methods from extremes has been absent until relatively recently with the papers such as [@Voitalov_2019] and [Lee]. Forgoing the use of extreme value methods essentially downplays how deviations from the power law in the largest degrees may affect the network given that they are often much more influential than the vertices with smaller degrees.

Noting this, many have now started to use these extreme value methods when analysing networks. [@wan19] showed that extreme value estimation methods prove more robust than existing parametric approaches especially when the data are corrupted or the model has been misspecified, demonstrating that extreme value methods can be a valuable tool when studying networks. [@wang2022random] investigates a PA model with a affine preference function (similar to the BA model) with the added behaviour of heterogeneous reciprocal edges, they show that the distribution of in-degrees and out-degrees produced by such a model is multivariate regularly varying and they also have hidden regular variation. [Lee] using methods from discrete extremes modelled the right tail of network degrees with a  discretised variation of the generalised Pareto distribution (GPD) usually dubbed the Integer GPD (IGPD) where $X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)$ is defined by the survival:

$$
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
$$
for $v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R$.

The study of degree distributions, whether using extremes or not, usually does not reveal any information about the preference function even if PA was the underlying generating mechanism of the network. Here we address this gap in the literature, more specifically investigating if given the degree distribution of a network assumed to come from the PA model can we directly infer the model parameters? 
We build upon the foundation provided by [@rudas07] deriving the limiting degree distribution in terms of the preference function and suggest a class of preference functions that result in realistic tails that are heavy but are lighter than implied by the power law.

## Structure of this paper

@sec-gpa looks at the theoretical limiting behaviour of a GPA model when $m=1$ according to the results from [@rudas07], going on to introduce a flexible class of preference functions that can guarantee a heavy tailed degree distribution while still being flexible in the body. @sec-rec shows that the preference function parameters can be fairly well recovered from the degree distributions of networks simulated from the GPA model using various functions of the class introduced in @sec-gpa. Having shown that for simulated data the model parameters are recoverable from the degree distribution alone, @sec-real builds upon this and attempts to fit the model to degree distributions of real networks; modelling not only the degree distribution itself but also the estimates for the parameters of the preference function assuming the networks evolved according to the GPA scheme. Ultimately, @sec-conc discusses the main results of the paper, pitfalls of this method and future work.

# PA model

@@

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
$$
We are specifically interested in how the tail heaviness ($\xi$ in the IGPD) is affected by the preference function $b$, [@shimura12] introduces a quantity that will help in determining what domain of attraction this discrete distribution belongs to, that is:

For a distribution $F$ with survival function $\bar F$ and some $n\in\mathbb Z^+$ let:

$$
\Omega(F,n) = \left(\log\displaystyle\frac{\bar F (n+1)}{\bar F (n+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (n)}{\bar F (n+1)}\right)^{-1}
$$

[@shimura12] then states that if $\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy tailed with $\bar F(n) \sim n^{-\alpha}$. Additionally, if $\lim_{n\rightarrow\infty} \Omega(F,n) = 0$ then the distribution is light tailed. This allows us to show the following

:::{#prp-omega}
If $b(k) \rightarrow \infty$ as $k\rightarrow \infty$ then
$$
\bar F(k) = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}
$$
and

$$
\lim_{k\rightarrow\infty}\Omega(F,n) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
$$
:::
See [appendix] for the details of the proof.

@@

We can show that in order for the degree distribution to be heavy tailed, the preference function must be asymptotically linear i.e. $\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0$.

Consider the definition of heavy tails:

\begin{align*}
\lim_{k\rightarrow\infty}\Omega(F,n) &= c>0\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} &= c\\
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] &= \lambda^*c\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{(k+1)-(k)} &= \lambda^*c\\
\end{align*}

by Stolz-Cesaro Theorem:

$$
\lim_{k\rightarrow\infty}\frac{b(k)}{k} = \lambda^*c > 0.
$$

Since $b$ must be asymptotically linear in order to have positive tail heaviness we propose the following preference function:

@@

# Preferential Attachment with flexible heavy tail {#sec-model}

@@

$$
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0.
\end{cases}
$${#eq-polysurv}

@@

$$
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
$$

which must be solved numerically for most parameter choices. Also, note that $\lambda>\beta$ since when $\lambda\le \beta$ the infinite sum no longer converges and instead goes to infinity.

@@

```{r, echo=FALSE}
source('funcs.R')
```

@@

The survival (@eq-polysurv) can be connected to the IGPD mentioned in [@sec-intro] by using Stirling's approximation of the gamma function to obtain:

$$
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
$$
and utilising the expression from @eq-polysurv:

@@

# Recovery {#sec-rec}

The goal of was to provide a possible preference function that is able to explain the growth of real networks with varied tail behaviour in their degree distributions. This section aims to show that the parameters of the model in @sec-model can be recovered from the degree distribution of a network simulated from it.

The procedure for recovering the parameters begins with simulating a network from the model with $N=100,000$ vertices and $m=1$ given some set of parameters $\pmb\theta = (\alpha, \beta, \varepsilon, k_0)$, obtaining the degree counts in the form of a vector of degrees $\pmb x = (1,2,\ldots,M)$ and the number of vertices with those degrees $\pmb n$ where $M$ is the maximum degree. However, when it comes to fitting this model to real data in the next section, small degrees are extremely influential and will often cause the model to have a bad fit even though the rest of the data looks similar to that of the simulations; truncating the data above some value, say $l<k_0$, should allow the model to be more effectively fitted. This accounts in some way for the fact that this model only applies to trees, and could not possibly capture the behaviour for nodes with small degrees. Using the p.m.f derived in [@rudas07], the likelihood given that $x_i \ge l$ for all $i =1,2,\ldots,M$ is:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = \left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{x_i\ge k_0}n_{x_i}\right)} \prod_{l \le x_i<k_0}\left(\frac{\lambda^*}{\lambda^* +x_i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ \times \prod_{x_i\ge k_0}\left(\frac{\text{B}(x_i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}

where $B(y,z)$ is the the beta function.

This likelihood allows inference to be made about the parameters using a Bayesian approach using the priors:

@@

:::{#fig-rec2 out.width="80%"}
![: Posterior estimates of paramters for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.](mcmc_plot.jpg)
:::

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. [describe networks being used]. Alongside fitting this model to the degree distributions, we then compare the fit to that of an mixture distribution that was used in [clement] and is similar to others that have been used to model degree distributions [others].

@@

:::{#fig-rec2 out.width="80%"}
![: Posterior estimates of paramters for real data.](pars_plot.png)
:::

# Conclusion and Discussion {#sec-conc}

@@

GPA is also a fairly simplistic one on its own despite having large flexibility in the preference function allowing for edges to only be added at times when a new node is introduced into the network and not allowing for anything like vertex/edge death, reciprocal edges and rewiring steps. [@deijfen2015] introduces some theory for GPA with vertex death based on the work by [@rudas07], but is unable to come to an expression for the degree distribution. However, adding something like fitness scores into the GPA model seems fairly simple to do even going so far as making each of the parameters used in the class of functions here node-wise instead of universal. For example having the preference weight for a vertex be given by $b(k_i, \underline{\gamma_i})$  where $\underline{\gamma_i}\sim G$ is the vertex's fitness and $k_i$ is the vertex's degree, it is reasonable to expect that the degree distribution for a model like this to be given by:

$$
\bar F(k) = \mathbb E_G\left[\prod_{i=0}^k \frac{b(i, \gamma)}{\lambda^* + b(i,\gamma)}\right],\qquad \gamma\sim G,\, k=0,1,2,\ldots
$$

If this were the case, the posterior distributions of parameters from @sec-real could be interpreted as the approximate distributions from which a vertex's fitness values are drawn instead of providing estimates of the universal parameter values. Additionally, it allows the ability to set some parameters as universal and some as vertex-wise for example one could let $\alpha, \beta, k_0$ be the same for all vertices but let each vertex in the network have its own "zero appeal" $\varepsilon_i\overset{\text{iid}}{\sim} \text{Ga}(\theta_1, \theta_2)$; moving the model to being more realistic and accounting for the fact that in many real networks the objects vertices represent have some value instrinsic to themselves other than the number of connections they have.

# Appendix {.appendix}

## Additional Results 
### Tail heaviness of general preferential attachment model

Recall the limiting survival function:

$$
\bar F(k) = \prod_{i=0}^k \frac{b(i)}{\lambda^*+b(i)},
$$

The aim is to determine how this distribution behaves for different choices of $b$, specifically what maximum domain of attraction does this distribution belong to and is it affected by the choice of preference function $b$. @shimura12 introduces a quantity that will help in determining what domain of attraction a discrete distribution belongs to, that is:

For a distribution $F$ with survival function $\bar F$ and some $n\in\mathbb Z^+$ let:

$$
\Omega(F,n) = \left(\log\displaystyle\frac{\bar F (n+1)}{\bar F (n+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (n)}{\bar F (n+1)}\right)^{-1}
$$

[@shimura12] then states that if $\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy tailed with $\bar F(n) \sim n^{-\alpha}$. Additionally, if $\lim_{n\rightarrow\infty} \Omega(F,n) = 0$ then the distribution is light tailed.

Consider a preference function $b(\cdot)$ that fulfills:

$$
\lim_{n\rightarrow \infty} b(n) = \infty
$${#eq-condb1}

Substituting in the form of $\bar F(n)$ from @eq-surv and taking the limit, subject to @eq-condb1:

$$
\lim_{n\rightarrow\infty}\Omega(F,n) = \lim_{n\rightarrow\infty}\frac{b(n+2)-b(n+1)}{\lambda^*}.
$$ {#eq-omegares} (see appendix for full proof).

Whilst there are many classes of functions that satisfy @eq-condb1 and @eq-condb2, the vast majority of these result in $\Omega(F,n) \rightarrow 0$, meaning that the limiting degree distribution is light tailed. In fact, in order for the limiting degree distribution to be heavy tailed, $b$ must be asymptotically linear ($b(k)\sim k$).

## Proofs

### Tail heaviness of GPA

@@

```{=tex}
\begin{align*}
\Omega(F,n)&=\left(\log\frac{\prod_{i=0}^{n+1}\frac{b(i)}{\lambda+b(i)}}{\prod_{i=0}^{n+2}\frac{b(i)}{\lambda+b(i)}}\right)^{-1}-\left(\log\frac{\prod_{i=0}^{n}\frac{b(i)}{\lambda+b(i)}}{\prod_{i=0}^{n+1}\frac{b(i)}{\lambda+b(i)}}\right)^{-1}\\
&=\left(\log\frac{\lambda+b(n+2)}{b(n+2)}\right)^{-1}-\left(\log\frac{\lambda+b(n+1)}{b(n+1)}\right)^{-1}\\
&=\left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}
\end{align*}
```
Clearly if $b(n)=c$ or $\lim_{n\rightarrow\infty}b(n)=c$ for some $c>0$ then $\Omega(F,n)=0$. Now consider a non-constant $b(n)$ and re-write $\Omega(F,n)$ as:

```{=tex}
\begin{align*}
\Omega(F,n) &= \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}+\frac{b(n+2)}{\lambda}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}+\frac{b(n+1)}{\lambda}  -\frac{b(n+1)}{\lambda}\\
&=\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}+\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}
\end{align*}
```
Then if $\lim_{n\rightarrow\infty}b(n)=\infty$ it follows that:

```{=tex}
\begin{align*}
\lim_{n\rightarrow\infty}\Omega(F,n) &= \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\
&\qquad+\lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{2}-\frac{1}{2} + \lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{\lambda}\lim_{n\rightarrow\infty}\left[b(n+2)-b(n+1)\right]\qquad \square
\end{align*}
```

### Removing the infinite sum

>>>>

---
title: Tail Flexibility in the Degrees of Preferential
 Attachment Networks
format:
  nature-pdf:
    # journal.cite-style is included in the tex file but ignored by pandoc if 
    # cite-method is not `natbib`.
    journal:
      cite-style: sn-basic
    # `citeproc` is the pandoc default. Set `cite-method: natbib` if required
    # to use the bst styles from the upstream template.
    cite-method: natbib
    keep-tex: true
    equal-margins: false
    number-sections: true
author:
  - name: Thomas William Boughen
    affiliations:
      - name: Newcastle University
        id: 1
        department: School of Mathematics, Statistics and Physics
  - name: Clement Lee
    affiliations:
      - ref: 1
  - name: Vianey Palacios Ramirez
    affiliations:
      - ref: 1
abstract: |
  Devising the underlying generating mechanism of a real-life network is difficult as, more often than not, only its snapshots are available, but not its full evolution. One candidate for the generating mechanism is preferential attachment which, in its simplest form, results in a degree distribution that follows the power law. Consequently, the growth of real-life networks that roughly display such power-law behaviour is commonly modelled by preferential attachment. However, the validity of the power law has been challenged by the presence of alternatives with comparable performance, as well as the recent findings that the right tail of the degree distribution is often lighter than implied by the body, whilst still being heavy. In this paper, we study a modified version of the model with a flexible preference function that allows super/sub-linear behaviour whilst also guaranteeing that the limiting degree distribution has a heavy tail. We relate the distributions tail heaviness directly to the model parameters, allowing direct inference of the parameters from the degree distribution alone.
keywords: [networks, extremes]
bibliography: refs.bib  
---

\newpage

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(paletteer)
library(gridExtra)
library(latex2exp)
library(ggpubr)
library(Rcpp)
library(RcppArmadillo)
```

# Introduction

Networks appear in many fields such as sociology, politics, epidemiology, and economics. Statistical methods have been and continue to be used to study networks that appear in these fields and beyond, from the use of stochastic block models to detect communities within the French political blogosphere [@Latouche11], to using Exponential Random Graph Models (ERGMs) to analyse the global trade network [@Setayesh22], and the use of mechanistic models in neuroscience to investigate the patterns of connections in neural systems [@Betzel17].

One example of a mechanistic model that may be considered is the preferential attachment (PA) model where as a network grows and new vertices join they connect to those already present in the network with weights proportional to some preference function $b(\cdot)$ of their in-degree. The well-known Barabási-Albert (BA) model is a special case of this when $b(k) = k+1$ [@Barabasi99].

These mechanistic models are generally useful when simulating networks, but not for fitting to real data as usually only snapshots of a network are available and not the full evolution. However, one statistic that is usually available for snapshots of networks is the degree distribution making it valuable to study the degrees in a network. Often, when looking at the degree distributions of real networks it can be attractive to use the power law when modelling as it is seemingly observed in much of the data. It also comes with the added benefit of offering a connection to preferential attachment since the BA model results in a power law degree distribution [@Barabasi99], perhaps suggesting that if the power law fits well then the network may have grown according to similar simple rules. However, attempting to explain the networks growth using preferential attachment when the degrees seemingly follow a power law may not be the best course of action as [@krapivsky01] shows that a general preference function does not lead to power law degrees and additionally, it is still being argued whether or not a lot of real data actually follows a power law. @Broido_2019 provides some evidence that suggests that real world networks rarely have degrees that follow a power law through the use of statistical analysis of almost 1000 networks across a wide variety of fields; @Voitalov_2019, on the other hand, disagrees and claims that these kinds of networks are not nearly as rare and only appear so as a result of an unrealistic expectation of power law without deviations or noise.

Extreme value methods have been scarcely used when it comes to analysis of degree distributions and whether they follow the power law. The absence of these methods essentially down weights the deviation of vertices with the largest degrees, which are usually the most influential, from the power law. Somewhat recent publications [@Voitalov_2019; @Lee24; @wang2022random] have addressed this gap in the literature. @Voitalov_2019 proposes estimating the power law exponent of a degree distribution using three estimators based in extreme value theory, in contrast to the estimators generally used in network science, the degree distributions are then split into one of three categories (not power-law , hardly power-law or power-law) depending on the values of the estimates obtained from each of the estimators. @Lee24 proposes a framework for modelling entire degree distributions using a combination of two generalisations of the power law including a model selection step that is capable of suggesting whether or not the data is adequately modelled by a power law, with one conclusion being that most data seems to have a tail lighter than what is implied by the body whilst still being heavy tailed. @wang2022random considers a preferential attachment model with an affine preference function and reciprocal edges, they study the limiting degree distribution of this model for both the in-degrees and out-degrees - finding that the joint limiting distribution is multivariate regularly varying and has the property of hidden regular variation.

Evidently, the study of degree distributions is a very active but the vast majority of research in the field (whether using extremes or not) are purely descriptive in the sense that no information about the preference function is revealed even if preferential attachment was the underlying mechanism. This paper aims to address this gap - asking if given the degree distribution of a network assumed to come from the preferential attachment model, can we directly infer the model parameters?

@rudas07 derives the limiting degree distribution in terms of the preference function $b(\cdot)$ and its parameters, we build upon this by finding a class of preference functions that result in realistic tails i.e. heavy but not as heavy as implied by the power law.

The remainder of this paper begins with a more detailed description of the PA model alongside the theoretical results for the limiting degree distribution using results from @rudas07, this section (@sec-tail) then goes on to analyse the tail behaviour of this limiting distribution in terms of the preference function $b(\cdot)$ arguing that a heavy tailed degree distribution can only be the result of an asymptotically linear preference function - ending with a suggestion for a preference function that can guarantee heavy tails while remaining flexible. @sec-model continues on from the previous section by considering the limiting distribution obtained when using a preference function of the form stated at the end of @sec-tail providing examples of how the shape of the distribution varies with respect to the model parameters as well as visualisations for how the tail heaviness changes when these parameters are varied - continued study of this distribution then also reveals a connection to the general Pareto distribution (GPD). @sec-sim consists of a simulation study where networks are simulated from the PA model for various parameter combinations and then the model specified in @sec-model is fitted to the resulting degree distributions using Bayesian inference, demonstrating that (for simulated data) the model parameters can be recovered from only the degree distribution. Finally, @sec-real fits the model to some real data and provides posterior estimates for the preference function assuming the networks evolved according to the PA model.

\newpage

# Tail Behaviour of Preferential Attachment Model {#sec-tail}

@@

$$
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
$$

We are specifically interested in how the tail heaviness ($\xi$ in the IGPD) is affected by the preference function $b$, [@shimura12] introduces a quantity that will help in determining what domain of attraction this discrete distribution belongs to, that is:

For a distribution $F$ with survival function $\bar F$ and some $k\in\mathbb Z^+$ let:

$$
\Omega(F,k) = \left(\log\displaystyle\frac{\bar F (k+1)}{\bar F (k+2)}\right)^{-1} - \left(\log\displaystyle\frac{\bar F (k)}{\bar F (k+1)}\right)^{-1}
$$

[@shimura12] then states that if $\lim_{n\rightarrow\infty} \Omega(F,k) = 1/\alpha$ ($\alpha>0$), then $F$ is heavy tailed with $\bar F(k) \sim k^{-\alpha}$. Additionally, if $\lim_{n\rightarrow\infty} \Omega(F,k) = 0$ then the distribution is light tailed. This allows us to show the following:

::: {#prp-omega}
If $b(k) \rightarrow \infty$ as $k\rightarrow \infty$ then $$
\bar F(k) = \prod_{i=0}^k\frac{b(i)}{\lambda^* + b(i)}
$$ and

$$
\lim_{k\rightarrow\infty}\Omega(F,k) = \lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*}.
$$
:::

See appendix [-@sec-proofs] for the details of the proof.

@@

We can show that in order for the degree distribution to be heavy tailed, the preference function must be asymptotically linear i.e. $\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0$. First we must consider the following theorem:

::: {#thm-stolz}
## Stolz-Cesàro Theorem [@cesaro]

Let $(a_n)_{n\ge1}$ and $(b_n)_{n\ge1}$ be two sequence of real numbers. Assume $(a_n)_{n\ge1}$ is a strictly monotone and divergent sequence and the following limit exists:

$$
\lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = l.
$$ Then,

$$
\lim_{k\rightarrow\infty}\frac{b_k}{a_k} = l.
$$
:::

From @prp-omega we know that in order for the limiting degree distribution to have heavy tails, we must have that:

$$
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = c>0.
$$ Now, setting $b_k = b(k)$ and $a_k = k$:

$$
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] = \lim_{k\rightarrow\infty}\frac{b_{k+1} - b_k}{a_{k+1} - a_k} = c>0,
$$ meaning by @thm-stolz:

$$
\lim_{k\rightarrow\infty}\frac{b_k}{a_k} = \lim_{k\rightarrow\infty}\frac{b(k)}{k} = c>0.
$$ Therefore, in order for the degree distribution to be heavy tailed, the preference function must be asymptotically linear. With this under consideration we propose the following preference function:

@@

\newpage

# Preferential Attachment with Flexible Heavy Tail {#sec-model}

@@

$$
\bar F(k) = \begin{cases}
\prod_{i=0}^{k}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon},&k<k_0\\
\left(\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda^*+i^\alpha + \varepsilon}\right)\frac{\Gamma(\lambda^*+k_0^\alpha + \varepsilon)/\beta)}{\Gamma\left((k_0^\alpha + \varepsilon)/\beta\right)} \frac{\Gamma\left(k-k_0 + 1 +\frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0 + 1 +\frac{\lambda^* +k_0^\alpha + \varepsilon}{\beta}\right)},&k\ge k_0.
\end{cases}
$$ {#eq-polysurv}

@@

$$
\hat\rho(\lambda) = \sum_{n=0}^{k_0}\prod_{i=0}^{n-1}\frac{i^\alpha + \varepsilon}{\lambda+i^\alpha + \varepsilon} + \left(\frac{k_0^\alpha + \varepsilon}{\lambda-\beta}\right)\prod_{i=0}^{k_0-1}\frac{i^\alpha + \varepsilon}{\lambda + i^\alpha + \varepsilon} 
$$ {#eq-rho}

which must be solved numerically for most parameter choices. Also, note that $\lambda>\beta$.

@@

```{r, echo=FALSE}
source('scripts/funcs.R')
```

@@

The survival (@eq-polysurv) can be connected to the IGPD by using Stirling's approximation of the gamma function, where the IGPD is a discretised variation of the generalised Pareto distribution (GPD) usually dubbed the Integer GPD (IGPD) where $X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)$ defined by the survival: $$
\Pr(X> x|X> v) = \left(\frac{\xi(x-v)}{\sigma} + 1\right)^{-1/\xi},\qquad x=v+1,v+2,\ldots
$$ for $v\in\mathbb Z^+, \sigma>0,\xi\in \mathbb R$.

Using Stirling's approximation we have that: $$
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
$$ now utilising the expression from @eq-polysurv:

@@

# Simulation Study {#sec-sim}

This section aims to show that the parameters of the model in @sec-model can be recovered from the degree distribution of a network simulated from it.

The procedure for recovering the parameters begins with simulating a network from the model with $N=100,000$ vertices and $m=1$ given some set of parameters $\pmb\theta = (\alpha, \beta, \varepsilon, k_0)$, obtaining the degree counts and using the likelihood from the previous section alongisde the priors:

@@

![: Posterior estimates of paramters for data simulated from the proposed model with various combinations of ($\alpha$,$\beta$,$\varepsilon$) and $k_0=20$.](images/mcmc_plot.jpg){#fig-rec2}

@@

Turning now to real data, the goal is to fit the model to the degree distributions of real networks from various sources. Alongside fitting this model to the degree distributions, we then compare the fit to that of an mixture distribution that was used in @Lee24.

@@

![: Posterior estimates of paramters for real data.](images/pars_plot.png)

# Conclusion and Discussion

@@

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

\newpage

# Proofs and Derivations {#sec-proofs .appendix}

## Proof of @prp-omega

@@

\begin{align*}
\Omega(F,n)&=\left(\log\frac{\prod_{i=0}^{n+1}\frac{b(i)}{\lambda+b(i)}}{\prod_{i=0}^{n+2}\frac{b(i)}{\lambda+b(i)}}\right)^{-1}-\left(\log\frac{\prod_{i=0}^{n}\frac{b(i)}{\lambda+b(i)}}{\prod_{i=0}^{n+1}\frac{b(i)}{\lambda+b(i)}}\right)^{-1}\\
&=\left(\log\frac{\lambda+b(n+2)}{b(n+2)}\right)^{-1}-\left(\log\frac{\lambda+b(n+1)}{b(n+1)}\right)^{-1}\\
&=\left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}
\end{align*}

Clearly if $b(n)=c$ or $\lim_{n\rightarrow\infty}b(n)=c$ for some $c>0$ then $\Omega(F,n)=0$. Now consider a non-constant $b(n)$ and re-write $\Omega(F,n)$ as:

\begin{align*}
\Omega(F,n) &= \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}+\frac{b(n+2)}{\lambda}-\left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}\\&+\frac{b(n+1)}{\lambda}  -\frac{b(n+1)}{\lambda}\\
&=\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} - \left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\&+\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}
\end{align*}

Then if $\lim_{n\rightarrow\infty}b(n)=\infty$ it follows that:

\begin{align*}
\lim_{n\rightarrow\infty}\Omega(F,n) &= \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+2)}\right]\right)^{-1}-\frac{b(n+2)}{\lambda}\right\} \\&- \lim_{n\rightarrow\infty}\left\{ \left(\log\left[1+\frac{\lambda}{b(n+1)}\right]\right)^{-1}-\frac{b(n+1)}{\lambda}\right\}\\
&\qquad+\lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{2}-\frac{1}{2} + \lim_{n\rightarrow\infty}\left(\frac{b(n+2)}{\lambda}-\frac{b(n+1)}{\lambda}\right)\\
&=\frac{1}{\lambda}\lim_{n\rightarrow\infty}\left[b(n+2)-b(n+1)\right]\qquad \square
\end{align*}

## Derivation of @eq-rho
